---
title: "Reproducible Research: Peer Assessment 1"
author: "Jeff Shilling"
date: "Monday, November 10, 2014"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
The first step is to read the csv file.  Then calculate the total number of steps for each day.
```{r read_data, echo=TRUE}
activity = read.csv("activity.csv")
daily = aggregate(activity$steps, by=list(activity$date), FUN=sum)
```

## What is mean total number of steps taken per day?  
This histogram shows the distribution of total daily steps.  
```{r total_daily_steps, echo=FALSE}
meansteps = mean(daily$x, na.rm=T)
medsteps = median(daily$x, na.rm=T)

hist(daily$x, breaks = 25, 
     col = "lightblue", border = "black",
     main = "Total number of steps taken each day",
     xlab = "# Steps taken in a day", ylab = "Frequency")
abline(v = meansteps, col = "blue", lwd = 2, lty = 2)
```
  
The mean number of steps take each day is `r format(meansteps, nsmall=2, big.mark=",")`, shown as the blue dashed line in the histogram above.  The median number of steps is a quite similar `r format(medsteps, big.mark=",")`.  


## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
